---
layout: page
title: R for reproducible scientific analysis
subtitle: Introduction to R and RStudio
minutes: 45
---

```{r, include=FALSE}
source("tools/chunk-options.R")
```

> ### Learning Objectives {.objectives}
>
> * To gain familiarity with the various panes in the RStudio IDE
> * To gain familiarity with the buttons, short cuts and options in the Rstudio IDE
> * To understand variables and how to assign to them
> * To be able to manage your workspace in an interactive R session
> * To be able to use mathematical and comparison operations
> * To be able to call functions
>

### Introduction to RStudio

Welcome to the R portion of the Software Carpentry workshop. 

Throughout this lesson, we're going to teach you some of the fundamentals of
the R language as well as some best practices for organising code for
scientific projects that will make your life easier.

We'll be using RStudio: a free, open source R integrated development
environment. It provides a built in editor, works on all platforms (including
on servers) and provides many advantages such as integration with version
control and project management.

**Basic layout**

When you first open RStudio, you will be greeted by three panels:

  * The interactive R console (entire left)
  * Workspace/History (tabbed in upper right)
  * Files/Plots/Packages/Help (tabbed in lower right)

Once you open files, such as R scripts, a scripting panel will also open in the top left.

#### Work flow within Rstudio
There are two main ways one can work within Rstudio.

1. Test and play within the interactive R console then copy code into a .R file to run later.
  1.  This works well when doing small tests and initially starting off.
  2.  Becomes laboursome.
2. Start writing in an .R file and use Rstudio's command / short cut to push current line, selected lines or modified lines to the interactive R console.
  1. This is great way to start and work as all workings are saved for latter reference and can be read latter.

> #### Tip: Pushing to the interactive R console {.callout}
> To run the current line click on the `Run` button just above the file pane. Or use the short cut which can be see 
> by hovering the mouse over the button.
>
> To run a block of code, select it and then `Run`. If you have modified a line
> of code within a block of code you have just run. There is no need to reselct the section and `Run`, 
> you can use the next button along, `Re-run the previous region`. This will run the previous code block inculding 
> the modifications you have made.
>

### Introduction to R

A lot of your time in R will be spent in the R interactive console. This is where you
will run all of your code, and can be a useful environment to try out ideas before 
adding them to an R script file. This console in RStudio is the same as the one you
would get if you just typed in `R` in your commandline environment.

The first thing you will see in the R interactive session is a bunch of information, 
followed by a ">" and a blinking cursor. In many ways this is similar to the shell
environment you learnt about during the shell lessons: it operates on the same idea
of a "Read, evaluate, print loop": you type in commands, R tries to execute them, and 
then returns a result.

#### Using R as a calculator

The simplest thing you could do with R is do arithmetic:

```{r}
1 + 100
```

And R will print out the answer, with a preceding "[1]". Don't worry about this 
for now, we'll explain that later. For now think of it as indicating ouput.

Just like bash, if you type in an incomplete command, R will wait for you to
complete it:

~~~ {.r}
> 1 +
~~~

~~~ {.output}
+
~~~

Any time you hit return and the R session shows a "+" instead of a ">", it 
means it's waiting for you to complete the command. If you want to cancel
a command you can simply hit "Esc" and RStudio will give you back the ">"
prompt.

> #### Tip: Cancelling commands {.callout}
>
> If you're using R from the commandline instead of from within RStudio,
> you need to use `Ctrl+C` instead of `Esc` to cancel the command. This
> applies to Mac users as well!
>
> Cancelling a command isn't just useful for killing incomplete commands:
> you can also use it to tell R to stop running code (for example if its
> taking much longer than you expect), or to get rid of the code you're 
> currently writing.
>

When using R as a calculator, the order of operations is the same as you
would have learnt back in school.

From highest to lowest precedence:

 * Brackets: `(`, `)`
 * Exponents: `^` or `**`
 * Divide: `/`
 * Multiply: `*`
 * Add: `+`
 * Subtract: `-`

```{r}
3 + 5 * 2
```

Use brackets (actually parentheses) to group to force the order of evaluation
if it differs from the default, or to set your own order.

```{r}
(3 + 5) * 2
```

But this can get unwieldy when not needed:

```{r, eval=FALSE}
(3 + (5 * (2 ^ 2))) # hard to read
3 + 5 * 2 ^ 2       # easier to read, once you know rules
3 + 5 * (2 ^ 2)     # if you forget some rules, this might help
``` 


The text I've typed after each line of code is called a comment. Anything that
follows on from the octothorpe (or hash) symbol `#` is ignored by R when it 
executes code.

Really small or large numbers get a scientific notation:

```{r}
2/10000
``` 

Which is shorthand for "multiplied by `10^XX`". So `2e-4`
is shorthand for `2 * 10^(-4)`.

You can write numbers in scientific notation too:

```{r}
5e3  # Note the lack of minus here
```

#### Mathematical functions

R has many built in mathematical functions. To call a function,
we simply type its name, follow by and open and closing bracket.
Anything we type inside those brackets is called the function's
arguments:

```{r}
sin(1)  # trigonometry functions
```

```{r}
log(1)  # natural logarithm
```

```{r}
log10(10) # base-10 logarithm
```

```{r}
exp(0.5) # e^(1/2)
```

Don't worry about trying to remember every function in R. You
can simply look them up on google, or if you can remember the
start of the function's name, use the tab completion in RStudio.

This is one advantage that RStudio has over R on its own, it
has autocompletion abilities that allow you to more easily 
look up functions, their arguments, and the values that they
take.

Typing a `?` before the name of a command will open the help page
for that command. As well as providing a detailed description of
the command and how it works, scrolling ot the bottom of the 
help page will usually show a collection of code examples which
illustrate command usage. We'll go through an example later.

#### Comparing things

We can also do comparison in R:

```{r}
1 == 1  # equality (note two equals signs, read as "is equal to")
```

```{r}
1 != 2  # inequality (read as "is not equal to")
```

```{r}
1 <  2  # less than
```

```{r}
1 <= 1  # less than or equal to
```

```{r}
1 > 0  # greater than
```

```{r}
1 >= -9 # greater than or equal to
```

> #### Tip: Comparing Numbers {.callout}
> 
> A word of warning about comparing numbers: you should 
> never use `==` to compare two numbers unless they are
> integers (a data type which can specifically represent
> only whole numbers). 
>
> Computers may only represent decimal numbers with a 
> certain degree of precision, so two numbers which look
> the same when printed out by R, may actually have
> different underlying representations and therefore be 
> different by a small margin of error (called Machine 
> numeric tolerance). 
> 
> Instead you should use the `all.equal` function.
>
> Further reading: [http://floating-point-gui.de/](http://floating-point-gui.de/)
>

#### Variables and assignment

We can store values in variables using the assignment operator `<-`, like this:

```{r}
x <- 1/40
```

Notice that assignment does not print a value. Instead, we stored it for later
in something called a **variable**. `x` now contains the **value** `0.025`:

```{r}
x
```

More precisely, the stored value is a *decimal approximation* of
this fraction called a [floating point number](http://en.wikipedia.org/wiki/Floating_point).

Look for the `Environment` tab in one of the panes of RStudio, and you will see that `x` and its value
have appeared. Our variable `x` can be used in place of a number in any calculation that expects a number:

```{r}
log(x)
```

Notice also that variables can be reassigned:

```{r}
x <- 100
```

`x` used to contain the value 0.025 and and now it has the value 100.

Assignment values can contain the variable being assigned to:

```{r}
x <- x + 1 #notice how RStudio updates its description of x on the top right tab
```

The right hand side of the assignment can be any valid R expression.
The right hand side is *fully evaluated* before the assignment occurs.

Variable names can contain letters, numbers, underscores and periods. They
cannot start with a number nor contain spaces at all. Different people use
different conventions for long variable names, these include

  * periods.between.words
  * underscores\_between_words
  * camelCaseToSeparateWords

What you use is up to you, but **be consistent**. 

It is also possible to use the `=` operator for assignment:

```{r}
x = 1/40
```

But this is much less common among R users.  The most important thing is to
**be consistent** with the operator you use. There are occasionally places
where it is less confusing to use `<-` than `=`, and it is the most common
symbol used in the community. So the recommendation is to use `<-`.

#### Managing your environment

There are a few useful commands you can use to interact with the R session.

`ls` will list all of the variables and functions stored in the global environment
(your working R session):

```{r}
ls()
```

> #### Tip: hidden objects {.callout}
> 
> Just like in the shell, `ls` will hide any variables or functions starting
> with a "." by default. To list all objects, type `ls(all.names=TRUE)` 
> instead
>

Note here that we didn't given any arguments to `ls`, but we still needed to give
the brackets to tell R to call the function.

If we type `ls` by itself, R will print out the source code for that function!

```{r}
ls
```

You can use `rm` to delete objects you no longer need:

```{r}
rm(x)
```

If you have lots of things in your environment and want to delete all of them,
you can pass the results of `ls` to the `rm` function:

```{r}
rm(list = ls())
```

In this case we've combined the two. Just like the order of operations, anything
inside the innermost brackets is evaluated first, and so on. 

In this case we've specified that the results of `ls` should be used for the 
`list` argument in `rm`. When assigning values to arguments by name, you *must* 
use the `=` operator!! 

If instead we use `<-`, there will be unintended side effects, or you may just 
get an error message:

```{r, error=TRUE}
rm(list <- ls())
```

> #### Tip: Warnings vs. Errors {.callout} 
>
> Pay attention when R does something unexpected! Errors, like above,
> are thrown when R cannot proceed with a calculation. Warnings on the
> other hand usually mean that the function has run, but it probably
> hasn't worked as expected.
>
> In both cases, the message that R prints out usually give you clues
> how to fix a problem.
>


> #### Challenge 1 {.challenge}
>
> Draw diagrams showing what variables refer to what values after each 
> statement in the following program:
>
> ~~~ {.r}
> mass <- 47.5
> age <- 122
> mass <- mass * 2.3
> age <- age - 20
> ~~~
>

> #### Challenge 2 {.challenge}
>
> Run the code from the previous challenge, and write a command to
> compare mass to age. Is mass larger than age?
>

> #### Challenge 3 {.challenge}
>
> Clean up your working environment by deleting the mass and age
> variables.
>

