---
layout: page
title: R for reproducible scientific analysis
subtitle: Data frames and reading in data
minutes: 45
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
gapminder <- read.csv("data/gapminder-FiveYearData.csv", header=TRUE)
```

> ## Learning Objectives {.objectives}
>
> * Become familiar with data frames
> * To be able to read in regular data into R
>

### Data frames

Data frames are similar to matrices, except each column can be a different atomic type.
Underneath the hood, data frames are really lists, where each element is
an atomic vector, with the added restriction that they're all the same length.
As you will see, if we pull out one column of a data frame, we will have a vector.


Data frames can be created manually with the `data.frame` function:

```{r}
df <- data.frame(id = c('a', 'b', 'c', 'd', 'e', 'f'), x = 1:6, y = c(214:219))
df
```

> #### Challenge: Dataframes {.challenge}
>
> Try using the `length` function to query
> your dataframe `df`. Does it give the result
> you expect?
>

Each column in the data frame is simply a list element, which is why when you ask for the
`length` of the data frame, it tells you the number of columns. If you actually want
the number of rows, you can use the `nrow` function.

We can add rows or columns to a data.frame using `rbind` or `cbind` (these are
the two-dimensional equivalents of the `c` function):

```{r}
df <- rbind(df, list("g", 11, 42))
```

This doesn't work as expected! What does this error message tell us?

It sounds like it was trying to generate a factor level. Why? Perhaps our first
column (containing characters) is to blame...
We can access a column in a `data.frame` by using the `$` operator.


```{r}
class(df$id)
```

Indeed, R automatically made this first column a factor, not a character vector.
We can change this in place by converting the type of this column.

```{r}
df$id <- as.character(df$id)
class(df$id)
```

Okay, now let's try adding that row again.

```{r}
df <- rbind(df, list("g", 11, 42))
tail(df, n=3)
```

Note that to add a row, we need to use a list, because each column is a different type!
If you want to add multiple rows to a data.frame, you will need to separate the new columns
in the list:

```{r}
df <- rbind(df, list(c("l", "m"), c(12, 13), c(534, -20)))
tail(df, n=3)
```

You can also row-bind data.frames together:

```{r}
rbind(df, df)
```

To add a column we can use `cbind`:

```{r}
df <- cbind(df, 10:1)
df
```

> #### Challenge {.challenge}
>
> Create a dataframe that holds the following information for yourself:
>
> * First name
> * Last name
> * Age
>
> Then use rbind to add the same information for the people sitting near you.
>
> Now use cbind to add a column of logicals answering the question,
> "Is there anything in this workshop you're finding confusing?"
>

### Reading in data

Remember earlier we obtained the gapminder dataset.
If you're curious about where this data comes from you might like to
look at the [Gapminder website](http://www.gapminder.org/data/documentation/)

Now we want to load the gapminder data into R.
Before reading in data, it's a good idea to have a look at its structure.
Let's take a look using our newly-learned shell skills:

```{r, engine='bash'}
cd data/ # navigate to the data directory of the project folder
head gapminder-FiveYearData.csv
```

As its file extension would suggest, the file contains comma-separated
values, and seems to contain a header row.

We can use `read.table` to read this into R

```{r}
gapminder <- read.table(
  file="data/gapminder-FiveYearData.csv",
  header=TRUE, sep=","
)
head(gapminder)
```

Because we know the structure of the data, we're able
to specify the appropriate arguments to `read.table`.
Without these arguments, `read.table` will do its
best to do something sensible, but it is always more
reliable to explicitly tell `read.table` the structure
of the data. `read.csv` function provides a convenient shortcut
for loading in CSV files.

```{r}
# Here is the read.csv version of the above command
gapminder <- read.csv(
  file="data/gapminder-FiveYearData.csv"
)
head(gapminder)
```

> #### Miscellaneous Tips {.callout}
>
> 1. Another type of file you might encounter are tab-separated
> format. To specify a tab as a separator, use `"\t"`.
>
> 2. You can also read in files from the internet by replacing
> the file paths with a web address.
>
> 3. You can read directly from excel spreadsheets without
> converting them to plain text first by using the `xlsx` package.
>

To make sure our analysis is reproducible, we should put the code
into a script file so we can come back to it later.

> #### Challenge {.challenge}
>
> Go to file -> new file -> R script, and write an R script
> to load in the gapminder dataset. Put it in the `scripts/`
> directory and add it to version control.
>
> Run the script using the `source` function, using the file path
> as its argument (or by pressing the "source" button in RStudio).
>

### Using dataframes: the `gapminder` dataset


To recap what we've just learnt, let's have a look at our example
data (life expectancy in various countries for various years).

Remember, there are a few functions we can use to interrogate data structures in R:

```{r, eval=FALSE}
class() # what is the data structure?
typeof() # what is its atomic type?
length() # how long is it? What about two dimensional objects?
attributes() # does it have any metadata?
str() # A full summary of the entire object
dim() # Dimensions of the object - also try nrow(), ncol()
```

Let's use them to explore the gapminder dataset.

```{r}
typeof(gapminder)
```

Remember, data frames are lists 'under the hood'.

```{r}
class(gapminder)
```

The gapminder data is stored in a "data.frame". This is the default data structure when you read
in data, and (as we've heard) is useful for storing data with mixed types of columns.

Let's look at some of the columns.

> #### Challenge: Data types in a real dataset {.challenge}
>
> Look at the first 6 rows of the gapminder dataframe we loaded before:
>
> ~~~ {.r}
> head(gapminder)
> ~~~
>
> ~~~ {.output}
> ##       country year      pop continent lifeExp gdpPercap
> ## 1 Afghanistan 1952  8425333      Asia  28.801  779.4453
> ## 2 Afghanistan 1957  9240934      Asia  30.332  820.8530
> ## 3 Afghanistan 1962 10267083      Asia  31.997  853.1007
> ## 4 Afghanistan 1967 11537966      Asia  34.020  836.1971
> ## 5 Afghanistan 1972 13079460      Asia  36.088  739.9811
> ## 6 Afghanistan 1977 14880372      Asia  38.438  786.1134
> ~~~
>
> Write down what data type you think is in each column
>


```{r}
typeof(gapminder$year)
typeof(gapminder$lifeExp)
```

Can anyone guess what we should expect the type of the continent column to be?

```{r}
typeof(gapminder$continent)
```

If you were expecting a the answer to be "character", you would rightly be
surprised by the answer. Let's take a look at the class of this column:

```{r}
class(gapminder$continent)
```

One of the default behaviours of R is to treat any text columns as "factors"
when reading in data. The reason for this is that text columns often represent
categorical data, which need to be factors to be handled appropriately by
the statistical modelling functions in R.

However it's not obvious behaviour, and something that trips many people up. We can
disable this behaviour and read in the data again.

```{r}
options(stringsAsFactors=FALSE)
gapminder <- read.table(
  file="data/gapminder-FiveYearData.csv", header=TRUE, sep=","
)
```

Remember, if you do turn it off automatic conversion to factors you will need to
explicitly convert the variables into factors when
running statistical models.
This can be useful, because it forces you to think about the
question you're asking, and makes it easier to specify the ordering of the categories.

However there are many in the R community who find it more sensible to
leave this as the default behaviour.

> #### Tip: Changing options {.callout}
>
> When R starts, the first thing it does is runs any code in the file `.Rprofile`
> in the project directory. Any permanent changes to default behaviour you want
> to make should be stored in that file.
>

The first thing you should do when reading data in, is check that it matches what
you expect, even if the command ran without warnings or errors. The `str` function,
short for "structure", is really useful for this:

```{r}
str(gapminder)
```

We can see that the object is a `data.frame` with 1,704 observations (rows),
and 6 variables (columns). Below that, we see the name of each column, followed
by a ":", followed by the type of variable in that column, along with the first
few entries.

We can also retrieve or modify the column or rownames of the data.frame:

```{r}
colnames(gapminder)
```

```{r}
rownames(gapminder)
```

See those numbers in the square brackets on the left? That tells you the
number of the first entry in that row of output. So in the last line, we
see that the "[1701]" element has "1701" stored in it. The rownames in
this case are simply the row numbers.

We can also modify this information:

```{r}
copy <- gapminder # lets create a copy so we don't mess up the original
colnames(copy) <- c("a", "b", "c", "d", "e", "f")
head(copy)
```

There are a few related ways of retreiving and modifying this information.
`attributes` will give you both the row and column names, along with the
class information, while `dimnames` will give you just the rownames and
column names.

In both cases, the output object is stored in a `list`:

```{r}
str(dimnames(df))
```

## Understanding how lists are used in function output


Lets run a basic linear regression on the gapminder dataset:

```{r}
# What is the relationship between life expectancy and year?
l1 <- lm(lifeExp ~ year, data=gapminder)
```

We won't go into too much detail of what I just wrote, but briefly;
the `~` denotes a formula, which means treat the variable on the left of the
`~` as the left hand side of the equation (or response in this case), and
everything on the right as the right hand side. By telling the linear
model function to use the gapminder data frame, it knows to look for those
variable names as its columns.

Let's look at the output:

```{r}
l1
```

Not much there right? But if we look at the structure...

```{r}
str(l1)
```

There's a lot of stuff, stored in nested lists! This is why the structure
function is really useful, it allows you to see all the data available to
you returned by a function.

For now, we can look at the `summary`:

```{r}
summary(l1)
```

As you might expect, life expectancy has slowly been increasing over
time, so we see a significant positive association!
