---
layout: page
title: R for reproducible scientific analysis
subtitle: Control flow
minutes: 45
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
gapminder <- read.csv("data/gapminder-FiveYearData.csv", header=TRUE)
# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```

> ## Learning Objectives {.objectives}
>
> * Write conditional statements with `if` and `else`.
> * Write and understand `while` and `for` loops. 
>

Often when we're coding we want to control the flow of our actions. This can be done 
by setting actions to occur only if a condition or a set of conditions are met. 
Alternatively, we can also set an action to occur a particular number of times.

There are several ways you can control flow in R. 
For conditional statements, the most commonly used approaches are the constructs: 

```{r, eval=FALSE}
# if
if (condition is true) {
  perform action
}

# if ... else
if (condition is true) {
  perform action
} else {  # that is, if the condition is false,
  perform alternative action
}
```

Say, for example, that we want R to print a message if a variable `x` has a particular value:

```{r}
# sample a random number from a Poisson distribution
# with a mean (lambda) of 8

x <- rpois(1, lambda=8)

if (x >= 10) {
  print("x is greater than or equal to 10")
}
  
x
```

Note you may not get the same output as your neighbour because
you may be sampling different random numbers from the same distribution. 

Let's set a seed so that we all generate the same 'pseudo-random'
number, and then print more information:

```{r}
x <- rpois(1, lambda=8)

if (x >= 10) {
  print("x is greater than or equal to 10")
} else if (x > 5) {
  print("x is greater than 5")
} else {
  print("x is less than 5")
}
```

> #### Tip: pseudo-random numbers {.callout}
>
> In the above case, the function `rpois` generates a random number following a
> Poisson distribution with a mean (i.e. lambda) of 8. The function `set.seed`
> guarantees that all machines will generate the exact same 'pseudo-random' 
> number ([more about pseudo-random numbers](http://en.wikibooks.org/wiki/R_Programming/Random_Number_Generation)).
> So if we `set.seed(10)`, we see that `x` takes the value 8. You should get the 
> exact same number.
>

**Important:** when R evaluates the condition inside `if` statements, it is
looking for a logical element, i.e., `TRUE` or `FALSE`. This can cause some
headaches for beginners. For example:

```{r}
x  <-  4 == 3
if (x) {
  "4 equals 3"
}
```

As we can see, the message was not printed because the vector x is `FALSE` 

```{r}
x <- 4 == 3
x
```

> #### Challenge 1 {.challenge}
> 
> Use an `if` statement to print a suitable message
> reporting whether there are any records from 2002 in 
> the `gapminder` dataset.
> Now do the same for 2012.
>

Did anyone get a warning message like this?

```{r, echo=FALSE}
if (gapminder$year == 2012) {}
```

If your condition evaluates to a vector with more than one logical element,
the function `if` will still run, but will only evaluate the condition in the first
element. Here you need to make sure your condition is of length 1.

> #### Tip: `any` and `all` {.callout}
> The `any` function will return TRUE if at least one
> TRUE value is found within a vector, otherwise it will return `FALSE`. 
> This can be used in a similar way to the `%in%` operator.
> The function `all`, as the name suggests, will only return `TRUE` if all values in
> the vector are `TRUE`. 
>


### Repeating operations

Sometimes you will find yourself needing to repeat an operation until a certain
condition is met. You can do this with a `while` loop. 

```{r, eval=FALSE}
while(this condition is true){
  do a thing
} 
```

Let's try an example, shall we?  We'll try to come up with some simple code
that generates random numbers from a uniform distribution (the `runif` function)
between 0 and 1 until it gets one that's less than 0.1.  

```{r, error=TRUE}
while(z > 0.1){
  z <- runif(1)
  print(z)
}
```

But wait, that doesn't work!  What's the problem?

The problem is that we haven't defined `z`, and so the very first time the
while loop's condition is checked (`z > 0.1`), `while` just says "Okay, that's
not true so I'm not going to execute this block of code".  The same thing would
have happened if we defined `z` to be anything less than 0.1.  Let's fix it.

```{r}
z <- 1
while(z > 0.1){
  z <- runif(1)
  print(z)
}
```

> #### Challenge 2 {.challenge}
> 
> Use a `while` loop to construct a vector called 'pet_list'
> with the value:
> 'cat', 'dog', 'dog', 'dog', 'dog'
> (N.B. using a loop may not be the most efficient way
> to do this, but it illustrates the principle!)
>

`while` loops will not always be appropriate. If you want to iterate over
a set of values, when the order of iteration is important, and perform the 
same operation on each, a `for` loop will do the job. 
We saw `for` loops in the shell lessons earlier. This is the most 
flexible of looping operations, but therefore also the hardest to use 
correctly. Avoid using `for` loops unless the order of iteration is important: 
i.e. the calculation at each iteration depends on the results of previous iterations.

The basic structure of a `for` loop is:

```{r, eval=FALSE}
for(iterator in set of values){
  do a thing
}
```

For example:

```{r}
for(i in 1:10){
  print(i)
}
``` 

The `1:10` bit creates a vector on the fly; you can iterate
over any other vector as well.

We can use a `for` loop nested within another `for` loop to iterate over two things at
once.

```{r}
for (i in 1:5){
  for(j in c('a', 'b', 'c', 'd', 'e')){
    print(paste(i,j))
  }
}
``` 

Rather than printing the results, we could write the loop output to a new object.

```{r}
output_vector <- c() 
for (i in 1:5){
  for(j in c('a', 'b', 'c', 'd', 'e')){
    temp_output <- paste(i, j)
    output_vector <- c(output_vector, temp_output)
  }
}
output_vector
```

This approach can be useful, but 'growing your results' (building
the result object incrementally) is computationally inefficient, so avoid
it when you are iterating through a lot of values.

> #### Tip: don't grow your results {.callout}
>
> One of the biggest things that trips up novices and
> experienced R users alike, is building a results object
> (vector, list, matrix, data frame) as your for loop progresses.
> Computers are very bad at handling this, so your calculations
> can very quickly slow to a crawl. It's much better to define
> an empty results object before hand of the appropriate dimensions.
> So if you know the end result will be stored in a matrix like above,
> create an empty matrix with 5 row and 5 columns, then at each iteration
> store the results in the appropriate location.
>

A better way is to define your (empty) output object before filling in the values.
For this example, it looks more involved, but is still more efficient.

```{r}
output_matrix <- matrix(nrow=5, ncol=5)
j_vector <- c('a', 'b', 'c', 'd', 'e')
for (i in 1:5){
  for(j in 1:5){
    temp_j_value <- j_vector[j]
    temp_output <- paste(i, temp_j_value)
    output_matrix[i, j] <- temp_output
  }
}
output_vector2 <- as.vector(output_matrix)
output_vector2
``` 

> #### Challenge 3 {.challenge}
>
> Compare the objects output_vector and 
> output_vector2. Are they the same? If not, why not?
> How would you change the last block of code to make output_vector2
> the same as output_vector?
>

> #### Challenge 4 {.challenge}
>
> Write a script that loops through the `gapminder` data by continent and prints out
> whether the mean life expectancy is smaller or larger than 50
> years.
>

> #### Challenge 5 {.challenge}
>
> Modify the script from Challenge 4 to also loop over each 
> country. This time print out whether the life expectancy is
> smaller than 50, between 50 and 70, or greater than 70.
>

> #### Challenge 6 - Advanced {.challenge}
>
> Write a script that loops over each country in the `gapminder` dataset,
> tests whether the country starts with a 'B', and graphs life expectancy 
> against time as a line graph if the mean life expectancy is under 50 years. 
>

