---
layout: page
title: R for reproducible scientific analysis
subtitle: Why use R?
minutes: xx
---

```{r, include=FALSE}
source("tools/chunk-options.R")
```

> ### Learning Objectives {.objectives}
>
> * To get a taste of R's powerful visualisation capabilities
> * To get a taste of R's powerful statistical analysis capabilities
> * To show how visualisation and analysis support each other
>

### Introduction to R

Welcome to the R portion of the Software Carpentry workshop. We're going to show you how R and RStudio can help you understand large data sets. We'll also guide your first steps towards using them effectively for your own work.

> #### Installation
> (d/l RStudio)
> (d/l gapminder -- can we include in this repo?}
> (what else?)
>

Open RStudio, and in [the prompt panel], enter these commands:

```{r}
gapminder <- read.csv("data/gapminder-FiveYearData.csv", header=TRUE, sep=',')
head(gapminder)
```

You should see something like this: XXX formatting

      country year      pop continent lifeExp gdpPercap
1 Afghanistan 1952  8425333      Asia  28.801  779.4453
2 Afghanistan 1957  9240934      Asia  30.332  820.8530
3 Afghanistan 1962 10267083      Asia  31.997  853.1007
4 Afghanistan 1967 11537966      Asia  34.020  836.1971
5 Afghanistan 1972 13079460      Asia  36.088  739.9811
6 Afghanistan 1977 14880372      Asia  38.438  786.1134

We have some data about countries, years, "pop", "lifeExp", and "gdpPercap". Let's see if we can get a better handle on it:

```{r}
library(ggplot2)
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) + geom_point()
```

In the lower right panel, you should see a graph that RStudio produced in response to your command. What can you tell about this data set from this initial graph?
