---
title: Creating Publication-Quality Graphics with ggplot2
teaching: 60
exercises: 20
questions:
- "How can I create publication-quality graphics in R?"
objectives:
- "To be able to use ggplot2 to generate publication quality graphics."
- "To apply geometry, aesthetic, and statistics layers to a ggplot plot."
- "To manipulate the aesthetics of a plot using different colors, shapes, and lines."
- "To improve data visualization through transforming scales and paneling by group."
- "To save a plot created with ggplot to disk."
keypoints:
- "Use `ggplot2` to create plots."
- "Think about graphics in layers: aesthetics, geometry, statistics, scale transformation, and grouping."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("08-")
# Silently load in the data so the rest of the lesson works
gapminder <- read.csv("data/gapminder_data.csv", header = TRUE)
```

Plotting our data is one of the best ways to
quickly explore it and the various relationships
between variables.

There are three main plotting systems in R,
the [base plotting system][base], the [lattice][lattice]
package, and the [ggplot2][ggplot2] package.

[base]: http://www.statmethods.net/graphs/index.html
[lattice]: http://www.statmethods.net/advgraphs/trellis.html
[ggplot2]: http://www.statmethods.net/advgraphs/ggplot2.html

Today we'll be learning about the ggplot2 package, because
it is the most effective for creating publication quality
graphics.

ggplot2 is built on the grammar of graphics, the idea that any plot can be
expressed from the same set of components: a **data** set, a
**coordinate system**, and a set of **geoms** -- the visual representation of data
points.

The key to understanding ggplot2 is thinking about a figure in layers.
This idea may be familiar to you if you have used image editing programs like Photoshop, Illustrator, or
Inkscape.

Let's start off with an example:

```{r lifeExp-vs-gdpPercap-scatter, message=FALSE}
library("ggplot2")
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

It is instructive to break down this line of code piece by piece. *Really* 
learning the fundamentals now will save hours of debugging later.
First, there are 2 interesting things to notice:

1. The axes on the plot are **gdpPerCap** and **lifeExp**
2. They came from `aes(x = gdpPercap, y = lifeExp)`

In previous lessons, we learned to interact with dataframes using this 
syntax: `gapminder$gdpPercap`. `ggplot` and other tidyverse packages modify 
the usual rules of R evaluation and save us some typing.

There are 2 more interesting things to notice:

3. `mapping = aes(...)` sets up aesthetics, which will be explained shortly.
4. `ggplot() + geom_point()` syntax is an example of the grammar of graphics.

The syntax in item 4. is **e1 + e2** where **e1** is a ggplot object or theme, and 
**e2** is a plot component. Plot components are the graphical layers, so there 
can be multiple components on any plot.


By itself, the call to `ggplot` isn't enough to draw a figure. Only the base
shows up, which is more like a blank canvas:

```{r lifeExp-vs-gdpPercap-scatter2, message=FALSE}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp))
```

If we return to the previous example, we can see a layer being added:
```{r lifeExp-vs-gdpPercap-scatter3, message=FALSE}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

We need to tell `ggplot` how we want to visually represent the data, which we
do by adding a new **geom** layer. In our example, we used `geom_point`, which
tells `ggplot` we want to visually represent the relationship between **x** and
**y** as a scatterplot of points:


Now we can do something more interesting:
```{r lifeExp-vs-gdpPercap-scatter4, message=FALSE}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point()
```

Finally, we can talk about `aes`. `aes` is short for "aesthetics," which modify how the 
plot looks and feels. In this case we told `ggplot` we
want to plot the "gdpPercap" column of the gapminder data frame on the x-axis, and
the "lifeExp" column on the y-axis, and color code by continent.

There are many ways to configure aesthetics, and we'll see m ore throughout the hour.

The important thing to know is there are 2 ways to use aesthetics.

1. First way: **Mapping**
  * This gives us insights about our data. The example above breaks points down by continent
  * Mapping  configurations are "inherited" by components. We gave `ggplot` aesthetics, and `geom_point` inherited them.
  
2. Second way: **By Component**
  * this applies settings to ALL items, not a subset.
  
Let's see an example: turn all the points blue:
```{r lifeExp-vs-gdpPercap-scatterblue, message=FALSE}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(color="blue")
```

**All** components like `geom_point` have configurable aesthetics, for **all** items.
This is why you don't need to type `aes` inside a component. Let's see what happens if
you use `aes`:

```{r lifeExp-vs-gdpPercap-scatterblue-bug, message=FALSE}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(color="blue"))
```

Sometimes this is what you want, sometime's it's a  mistake. If it is a mistake,
now you know how to debug it! The bug here is, "blue" is not a column inside 
gapminder.

> ## Challenge 1
>
> Modify the example so that the figure shows how life expectancy has
> changed over time:
>
> ```{r, eval=FALSE}
> ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) + geom_point()
> ```
>
> Hint: the gapminder dataset has a column called "year", which should appear
> on the x-axis.
>
> > ## Solution to challenge 1
> >
> > Here is one possible solution:
> >
> > ```{r ch1-sol}
> > ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp)) + geom_point()
> > ```
> >
> {: .solution}
{: .challenge}

>
> ## Challenge 2
>
> In the previous examples and challenge we've used the `aes` function to tell
> the scatterplot **geom** about the **x** and **y** locations of each point.
> Another *aesthetic* property we can modify is the point *color*. Modify the
> code from the previous challenge to **color** the points by the "continent"
> column. What trends do you see in the data? Are they what you expected?
>
> > ## Solution to challenge 2
> >
> > The solution presented below adds `color=continent` to the call of the `aes`
> > function. The general trend seems to indicate an increased life expectancy
> > over the years. On continents with stronger economies we find a longer life
> > expectancy.
> >
> > ```{r ch2-sol}
> > ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp, color=continent)) +
> >   geom_point()
> > ```
> >
> {: .solution}
{: .challenge}


## Layers

Using a scatterplot probably isn't the best for visualizing change over time.
Instead, let's tell `ggplot` to visualize the data as a line plot:

```{r lifeExp-line}
ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line()
```

Instead of adding a `geom_point` layer, we've added a `geom_line` layer. We've
added the **by** *aesthetic*, which tells `ggplot` to draw a line for each
country.

To restate what we've done, each line is a country, and each color is a continent.

But what if we want to visualize both lines and points on the plot? We can
add another layer to the plot:

```{r lifeExp-line-point}
ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line() + geom_point()
```

Does the order matter in `geom_line` and `geom_point`? Order does matter, because 
it indicates how  layers are organised. Since `geom_point` is **after**
`geom_line()`, the points are drawn on **on top** of the lines.

Let's illustrate the point about layer order:

```{r lifeExp-layer-example-1}
ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, by=country)) +
  geom_line(mapping = aes(color=continent)) + geom_point()
```


Notice that we moved `aes(color=continent)` out of the base and into the 
`geom_line` component. Now we can clearly see that the points are drawn on top of the
lines.

> ## Tip: Setting an aesthetic to a value instead of a mapping
>
> So far, we've seen how to use an aesthetic (such as **color**) as a *mapping* to a variable in the data. For example, when we use `geom_line(mapping = aes(color=continent))`, ggplot will give a different color to each continent. But what if we want to change the colour of all lines to blue? You may think that `geom_line(mapping = aes(color="blue"))` should work, but it doesn't. Since we don't want to create a mapping to a specific variable, we can move the color specification outside of the `aes()` function, like this: `geom_line(color="blue")`.
{: .callout}

> ## Challenge 3
>
> Switch the order of the point and line layers from the previous example. What
> happened?
>
> > ## Solution to challenge 3
> >
> > The lines now get drawn over the points!
> >
> > ```{r ch3-sol}
> > ggplot(data = gapminder, mapping = aes(x=year, y=lifeExp, by=country)) +
> >  geom_point() + geom_line(mapping = aes(color=continent))
> > ```
> >
> >
> {: .solution}
{: .challenge}

## Transformations and statistics

ggplot2 also makes it easy to overlay statistical models over the data. To
demonstrate we'll go back to our first example:

```{r lifeExp-vs-gdpPercap-scatter5, message=FALSE}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

Currently it's hard to see the relationship between the points due to some strong
outliers in GDP per capita. We can change the scale of units on the x axis using
the *scale* functions. These control the mapping between the data values and
visual values of an aesthetic. We can also modify the transparency of the
points, using the *alpha* function, which is especially helpful when you have
a large amount of data which is very clustered.

```{r axis-scale}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5) + scale_x_log10()
```

The `log10` function applied a transformation to the values of the gdpPercap
column before rendering them on the plot, so that each multiple of 10 now only
corresponds to an increase in 1 on the transformed scale, e.g. a GDP per capita
of 1,000 is now 3 on the x axis, a value of 10,000 corresponds to 4 on the x
axis and so on. This makes it easier to visualize the spread of data on the
x-axis.

> ## Tip Reminder: Setting an aesthetic to a value instead of a mapping
>
> Notice that we used `geom_point(alpha = 0.5)`. As the previous tip mentioned, using a setting outside of the `aes()` function will cause this value to be used for all points, which is what we want in this case. But just like any other aesthetic setting, *alpha* can also be mapped to a variable in the data. For example, we can give a different transparency to each continent with `geom_point(mapping = aes(alpha = continent))`.
{: .callout}

We can fit a simple relationship to the data by adding another layer,
`geom_smooth`:

```{r lm-fit}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm")
```

We can make the line thicker by *setting* the **size** aesthetic in the
`geom_smooth` layer:

```{r lm-fit2}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm", size=1.5)
```

There are two ways an *aesthetic* can be specified. Here we *set* the **size**
aesthetic by passing it as an argument to `geom_smooth`. Previously in the
lesson we've used the `aes` function to define a *mapping* between data
variables and their visual representation.

> ## Challenge 4a
>
> Modify the color and size of the points on the point layer in the previous
> example.
>
> Hint: do not use the `aes` function.
>
> > ## Solution to challenge 4a
> >
> > Here a possible solution:
> > Notice that the `color` argument is supplied outside of the `aes()` function.
> > This means that it applies to all data points on the graph and is not related to
> > a specific variable.
> >
> > ```{r ch4a-sol}
> > ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
> >  geom_point(size=3, color="orange") + scale_x_log10() +
> >  geom_smooth(method="lm", size=1.5)
> > ```
> > 
> {: .solution}
{: .challenge}


> ## Challenge 4b
>
> Modify your solution to Challenge 4a so that the
> points are now a different shape and are colored by continent with new
> trendlines.  Hint: The color argument can be used inside the aesthetic.
>
> > ## Solution to challenge 4b
> >
> > Here is a possible solution:
> > Notice that supplying the `color` argument inside the `aes()` functions enables you to
> > connect it to a certain variable. The `shape` argument, as you can see, modifies all
> > data points the same way (it is outside the `aes()` call) while the `color` argument which
> > is placed inside the `aes()` call modifies a point's color based on its continent value.
> >
> > ```{r ch4b-sol}
> > ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
> >  geom_point(size=3, shape=17) + scale_x_log10() +
> >  geom_smooth(method="lm", size=1.5)
> > ```
> > 
> {: .solution}
{: .challenge}


## Multi-panel figures

Earlier we visualized the change in life expectancy over time across all
countries in one plot. Alternatively, we can split this out over multiple panels
by adding a layer of **facet** panels.

> ## Tip
>
> We start by making a subset of data including only countries located
> in the Americas.  This includes 25 countries, which will begin to
> clutter the figure.  Note that we apply a "theme" definition to rotate
> the x-axis labels to maintain readability.  Nearly everything in
> ggplot2 is customizable.
{: .callout}

```{r facet}
americas <- gapminder[gapminder$continent == "Americas",]
ggplot(data = americas, mapping = aes(x = year, y = lifeExp)) +
  geom_line() +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 45))
```

The `facet_wrap` layer took a "formula" as its argument, denoted by the tilde
(~). This tells R to draw a panel for each unique value in the country column
of the gapminder dataset.

## Modifying text

To clean this figure up for a publication we need to change some of the text
elements. The x-axis is too cluttered, and the y axis should read
"Life expectancy", rather than the column name in the data frame.

We can do this by adding a couple of different layers. The **theme** layer
controls the axis text, and overall text size. Labels for the axes, plot
title and any legend can be set using the `labs` function. Legend titles
are set using the same names we used in the `aes` specification. Thus below
the color legend title is set using `color = "Continent"`, while the title
of a fill legend would be set using `fill = "MyTitle"`.

```{r theme}
ggplot(data = americas, mapping = aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  labs(
    x = "Year",              # x axis title
    y = "Life expectancy",   # y axis title
    title = "Figure 1",      # main title of figure
    color = "Continent"      # title of legend
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


## Exporting the plot

The `ggsave()` function allows you to export a plot created with ggplot. You can specify the dimension and resolution of your plot by adjusting the appropriate arguments (`width`, `height` and `dpi`) to create high quality graphics for publication. In order to save the plot from above, we first assign it to a variable `lifeExp_plot`, then tell `ggsave` to save that plot in `png` format to a directory called `results`. (Make sure you have a `results/` folder in your working directory.)

```{r directory-check, echo = FALSE}
if (!dir.exists("results")) {
  dir.create("results")
}
```

```{r save}
lifeExp_plot <- ggplot(data = americas, mapping = aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  labs(
    x = "Year",              # x axis title
    y = "Life expectancy",   # y axis title
    title = "Figure 1",      # main title of figure
    color = "Continent"      # title of legend
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave(filename = "results/lifeExp.png", plot = lifeExp_plot, width = 12, height = 10, dpi = 300, units = "cm")
```

There are two nice things about `ggsave`. First, it defaults to the last plot, so if you omit the `plot` argument it will automatically save the last plot you created with `ggplot`. Secondly, it tries to determine the format you want to save your plot in from the file extension you provide for the filename (for example `.png` or `.pdf`). If you need to, you can specify the format explicitly in the `device` argument.

This is a taste of what you can do with ggplot2. RStudio provides a
really useful [cheat sheet][cheat] of the different layers available, and more
extensive documentation is available on the [ggplot2 website][ggplot-doc].
Finally, if you have no idea how to change something, a quick Google search will
usually send you to a relevant question and answer on Stack Overflow with reusable
code to modify!

[cheat]: https://www.rstudio.org/links/data_visualization_cheat_sheet
[ggplot-doc]: https://ggplot2.tidyverse.org/reference/


> ## Challenge 5
>
> Generate boxplots to compare life expectancy between the different continents during the available years.
>
> Advanced:
>  - Rename y axis as Life Expectancy.
>  - Remove x axis labels.
>
> > ## Solution to Challenge 5
> >
> > Here a possible solution:
> > `xlab()` and `ylab()` set labels for the x and y axes, respectively
> > The axis title, text and ticks are attributes of the theme and must be modified within a `theme()` call.
> >
> > ```{r ch5-sol}
> > ggplot(data = gapminder, mapping = aes(x = continent, y = lifeExp, fill = continent)) +
> >  geom_boxplot() + facet_wrap(~year) +
> >  ylab("Life Expectancy") +
> >  theme(axis.title.x=element_blank(),
> >        axis.text.x = element_blank(),
> >        axis.ticks.x = element_blank())
> > ```
> > 
> {: .solution}
{: .challenge}
